---
title: "grepreaper Package - Core Functionality Testing & Demonstration"
author: "Package Developer"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: cosmo
    highlight: tango
---

```{r setup, include=FALSE, eval=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  error = FALSE,
  fig.width = 10,
  fig.height = 6
)

# Load required packages
library(data.table)
library(grepreaper)

# Set data.table options for better display
options(datatable.print.nrows = 10)
options(datatable.print.topn = 5)
```

## Package Overview

The `grepreaper` package provides efficient file reading and filtering capabilities using the power of `grep` at the command line. It's designed for:

- **Cross-platform compatibility** (Windows, macOS, Linux)
- **Efficient large file processing** using `grep` and `data.table`
- **Flexible metadata handling** (line numbers, filenames)
- **CRAN-ready structure** for worldwide distribution

## Core Functions Available

```{r function-check, eval=FALSE}
# Check all available functions
cat("Available functions in grepreaper package:\n")
cat("==========================================\n")
functions <- c("grep_read", "split.columns", "check_grep_availability", 
               "safe_system_call", "get_system_info", "build_grep_cmd", "is_binary_file")

for (func in functions) {
  if (exists(func)) {
    cat("PASS:", func, "\n")
  } else {
    cat("FAIL:", func, "NOT available\n")
  }
}
```

## System Information & Compatibility

```{r system-info, eval=FALSE}
# Get system information
sys_info <- get_system_info()
cat("System Information:\n")
cat("==================\n")
cat("OS:", sys_info$os, "\n")
cat("Release:", sys_info$release, "\n")
cat("Machine:", sys_info$machine, "\n")
cat("Grep Available:", sys_info$grep_available, "\n")

# Check grep availability
grep_status <- check_grep_availability()
cat("\nGrep Status:\n")
cat("============\n")
cat("Available:", grep_status$available, "\n")
if (!is.null(grep_status$path)) {
  cat("Path:", grep_status$path, "\n")
}
```

## Test Data Setup

```{r data-setup}
# Check if test data exists
data_files <- list.files("data", full.names = TRUE)
cat("Available test data files:\n")
cat("==========================\n")
for (file in data_files) {
  file_size <- file.size(file)
  cat(basename(file), "-", round(file_size/1024, 1), "KB\n")
}

# Use diamonds.csv for testing
test_file <- "data/diamonds.csv"
if (file.exists(test_file)) {
  cat("\nPASS: Test file available:", basename(test_file), "\n")
} else {
  cat("\nFAIL: Test file not found:", test_file, "\n")
}
```

## Core Functionality Tests

### Test 1: Basic File Reading (Single File)

```{r test-basic}
cat("=== Test 1: Basic File Reading ===\n")

# Read first 10 rows without any filtering
result <- grep_read(files = test_file, nrows = 10)
cat("Result shape:", nrow(result), "x", ncol(result), "\n")
cat("Columns:", paste(names(result), collapse = ", "), "\n")
cat("First few rows:\n")
print(head(result, 3))
```

### Test 2: Pattern Matching with Fixed Strings

```{r test-pattern}
cat("=== Test 2: Pattern Matching ===\n")

# Search for diamonds with "VS" clarity (fixed string)
vs_diamonds <- grep_read(files = test_file, pattern = "VS", fixed = TRUE, nrows = 5)
cat("Found", nrow(vs_diamonds), "diamonds with VS clarity\n")
cat("Sample results:\n")
print(head(vs_diamonds, 3))
```

### Test 3: Line Numbers (Single File)

```{r test-line-numbers-single}
cat("=== Test 3: Line Numbers (Single File) ===\n")

# Read with line numbers
result_with_lines <- grep_read(files = test_file, show_line_numbers = TRUE, nrows = 5)
cat("Result with line numbers:\n")
cat("Columns:", paste(names(result_with_lines), collapse = ", "), "\n")
print(head(result_with_lines, 3))

# Verify line numbers are sequential
if ("line_number" %in% names(result_with_lines)) {
  cat("Line numbers:", paste(result_with_lines$line_number, collapse = ", "), "\n")
  cat("Line numbers sequential:", all(diff(result_with_lines$line_number) == 1), "\n")
}
```

### Test 4: Filename Inclusion (Single File)

```{r test-filename-single}
cat("=== Test 4: Filename Inclusion (Single File) ===\n")

# Read with filename included
result_with_filename <- grep_read(files = test_file, include_filename = TRUE, nrows = 5)
cat("Result with filename:\n")
cat("Columns:", paste(names(result_with_filename), collapse = ", "), "\n")
print(head(result_with_filename, 3))

# Verify filename column
if ("source_file" %in% names(result_with_filename)) {
  cat("Source files:", unique(result_with_filename$source_file), "\n")
}
```

### Test 5: Multiple Files - No Metadata (CRITICAL TEST)

```{r test-multiple-no-metadata}
cat("=== Test 5: Multiple Files - No Metadata (CRITICAL TEST) ===\n")

# Test the scenario that was failing in previous testing
# Multiple files without filename or line numbers
multiple_files <- c(test_file, test_file)
result_multiple <- grep_read(
  files = multiple_files, 
  show_line_numbers = FALSE, 
  include_filename = FALSE,
  nrows = 10
)

cat("Multiple files result:\n")
cat("Shape:", nrow(result_multiple), "x", ncol(result_multiple), "\n")
cat("Columns:", paste(names(result_multiple), collapse = ", "), "\n")

# CRITICAL: Check if carat column has data (not NA)
if ("carat" %in% names(result_multiple)) {
  carat_na_count <- sum(is.na(result_multiple$carat))
  carat_total <- nrow(result_multiple)
  cat("Carat column - Total rows:", carat_total, "\n")
  cat("Carat column - NA rows:", carat_na_count, "\n")
  cat("Carat column - Data rows:", carat_total - carat_na_count, "\n")
  cat("Carat column - % NA:", round(carat_na_count/carat_total * 100, 1), "%\n")
  
  # Show sample carat values
  cat("Sample carat values:", paste(head(result_multiple$carat, 5), collapse = ", "), "\n")
}
```

### Test 6: Multiple Files - Line Numbers Only (CRITICAL TEST)

```{r test-multiple-line-numbers}
cat("=== Test 6: Multiple Files - Line Numbers Only (CRITICAL TEST) ===\n")

# Test multiple files with line numbers but no filename
result_multiple_lines <- grep_read(
  files = multiple_files, 
  show_line_numbers = TRUE, 
  include_filename = FALSE,
  nrows = 10
)

cat("Multiple files with line numbers:\n")
cat("Shape:", nrow(result_multiple_lines), "x", ncol(result_multiple_lines), "\n")
cat("Columns:", paste(names(result_multiple_lines), collapse = ", "), "\n")

# Check if data is preserved
if ("carat" %in% names(result_multiple_lines)) {
  carat_na_count <- sum(is.na(result_multiple_lines$carat))
  carat_total <- nrow(result_multiple_lines)
  cat("Carat column - % NA:", round(carat_na_count/carat_total * 100, 1), "%\n")
}
```

### Test 7: Multiple Files - Filename Only (Should Pass)

```{r test-multiple-filename}
cat("=== Test 7: Multiple Files - Filename Only (Should Pass) ===\n")

# Test multiple files with filename but no line numbers
result_multiple_filename <- grep_read(
  files = multiple_files, 
  show_line_numbers = FALSE, 
  include_filename = TRUE,
  nrows = 10
)

cat("Multiple files with filename:\n")
cat("Shape:", nrow(result_multiple_filename), "x", ncol(result_multiple_filename), "\n")
cat("Columns:", paste(names(result_multiple_filename), collapse = ", "), "\n")

# Check if data is preserved
if ("carat" %in% names(result_multiple_filename)) {
  carat_na_count <- sum(is.na(result_multiple_filename$carat))
  carat_total <- nrow(result_multiple_filename)
  cat("Carat column - % NA:", round(carat_na_count/carat_total * 100, 1), "%\n")
}
```

### Test 8: Multiple Files - Both Metadata (Should Pass)

```{r test-multiple-both}
cat("=== Test 8: Multiple Files - Both Metadata (Should Pass) ===\n")

# Test multiple files with both filename and line numbers
result_multiple_both <- grep_read(
  files = multiple_files, 
  show_line_numbers = TRUE, 
  include_filename = TRUE,
  nrows = 10
)

cat("Multiple files with both metadata:\n")
cat("Shape:", nrow(result_multiple_both), "x", ncol(result_multiple_both), "\n")
cat("Columns:", paste(names(result_multiple_both), collapse = ", "), "\n")

# Check if data is preserved
if ("carat" %in% names(result_multiple_both)) {
  carat_na_count <- sum(is.na(result_multiple_both$carat))
  carat_total <- nrow(result_multiple_both)
  cat("Carat column - % NA:", round(carat_na_count/carat_total * 100, 1), "%\n")
}
```

## Advanced Functionality Tests

### Test 9: Pattern Matching with Regular Expressions

```{r test-regex}
cat("=== Test 9: Regular Expression Pattern Matching ===\n")

# Search for diamonds with price starting with "3" (regex)
price_3_diamonds <- grep_read(files = test_file, pattern = "^3", nrows = 5)
cat("Found", nrow(price_3_diamonds), "diamonds with price starting with '3'\n")
cat("Sample results:\n")
print(head(price_3_diamonds, 3))
```

### Test 10: Count Only Mode

```{r test-count}
cat("=== Test 10: Count Only Mode ===\n")

# Count matches for different patterns
vs_count <- grep_read(files = test_file, pattern = "VS", fixed = TRUE, count_only = TRUE)
cat("VS clarity diamonds count:", vs_count$count, "\n")

# Count with filename
vs_count_with_file <- grep_read(files = multiple_files, pattern = "VS", fixed = TRUE, count_only = TRUE, include_filename = TRUE)
cat("VS clarity counts by file:\n")
print(vs_count_with_file)
```

### Test 11: Only Matching Mode

```{r test-only-matching}
cat("=== Test 11: Only Matching Mode ===\n")

# Get only the matching parts
vs_matches <- grep_read(files = test_file, pattern = "VS", fixed = TRUE, only_matching = TRUE, nrows = 10)
cat("Only matching parts (VS):\n")
print(head(vs_matches, 5))
```

### Test 12: Case Insensitive Search

```{r test-case-insensitive}
cat("=== Test 12: Case Insensitive Search ===\n")

# Search for "vs" (case insensitive)
vs_insensitive <- grep_read(files = test_file, pattern = "vs", ignore_case = TRUE, nrows = 5)
cat("Case insensitive 'vs' search - found", nrow(vs_insensitive), "rows\n")
```

### Test 13: Inverted Search

```{r test-inverted}
cat("=== Test 13: Inverted Search ===\n")

# Find rows that DON'T contain "VS"
not_vs <- grep_read(files = test_file, pattern = "VS", fixed = TRUE, invert = TRUE, nrows = 5)
cat("Rows NOT containing 'VS' - found", nrow(not_vs), "rows\n")
```

## Utility Function Tests

### Test 14: Split Columns Function

```{r test-split-columns}
cat("=== Test 14: Split Columns Function ===\n")

# Test the split.columns utility function
test_data <- c("file1:line1:data1", "file2:line2:data2", "file3:line3:data3")
split_result <- split.columns(test_data, c("file", "line", "data"), ":", 3)
cat("Split columns result:\n")
cat("Shape:", nrow(split_result), "x", ncol(split_result), "\n")
cat("Columns:", paste(names(split_result), collapse = ", "), "\n")
print(split_result)
```

### Test 15: Build Grep Command

```{r test-build-cmd}
cat("=== Test 15: Build Grep Command ===\n")

# Test command building
cmd <- build_grep_cmd(pattern = "VS", files = test_file, options = "-H -n")
cat("Built grep command:", cmd, "\n")
```

## Performance & Data Integrity Tests

### Test 16: Large Dataset Handling

```{r test-large-dataset}
cat("=== Test 16: Large Dataset Handling ===\n")

# Test with larger sample
large_sample <- grep_read(files = test_file, nrows = 100)
cat("Large sample result:\n")
cat("Shape:", nrow(large_sample), "x", ncol(large_sample), "\n")

# Check data types
cat("Column types:\n")
for (col in names(large_sample)) {
  cat(col, ":", class(large_sample[[col]])[1], "\n")
}
```

### Test 17: Empty Results Handling

```{r test-empty-results}
cat("=== Test 17: Empty Results Handling ===\n")

# Search for something that shouldn't exist
empty_result <- grep_read(files = test_file, pattern = "XYZ123", fixed = TRUE)
cat("Empty result handling:\n")
cat("Result type:", class(empty_result), "\n")
cat("Result shape:", nrow(empty_result), "x", ncol(empty_result), "\n")
```

## Summary of Critical Tests

```{r summary}
cat("=== CRITICAL TESTS SUMMARY ===\n")
cat("=============================\n")

# Test 5: Multiple files, no metadata
cat("Test 5 (Multiple files, no metadata): ")
if (exists("result_multiple") && "carat" %in% names(result_multiple)) {
  na_percent <- round(sum(is.na(result_multiple$carat))/nrow(result_multiple) * 100, 1)
  if (na_percent < 10) {
    cat("PASSED - Carat data preserved (", na_percent, "% NA)\n")
  } else {
    cat("FAILED - Carat data corrupted (", na_percent, "% NA)\n")
  }
} else {
  cat("FAILED - Test not completed\n")
}

# Test 6: Multiple files, line numbers only
cat("Test 6 (Multiple files, line numbers only): ")
if (exists("result_multiple_lines") && "carat" %in% names(result_multiple_lines)) {
  na_percent <- round(sum(is.na(result_multiple_lines$carat))/nrow(result_multiple_lines) * 100, 1)
  if (na_percent < 10) {
    cat("PASSED - Carat data preserved (", na_percent, "% NA)\n")
  } else {
    cat("FAILED - Carat data corrupted (", na_percent, "% NA)\n")
  }
} else {
  cat("FAILED - Test not completed\n")
}

# Test 7: Multiple files, filename only
cat("Test 7 (Multiple files, filename only): ")
if (exists("result_multiple_filename") && "carat" %in% names(result_multiple_filename)) {
  na_percent <- round(sum(is.na(result_multiple_filename$carat))/nrow(result_multiple_filename) * 100, 1)
  if (na_percent < 10) {
    cat("PASSED - Carat data preserved (", na_percent, "% NA)\n")
  } else {
    cat("FAILED - Carat data corrupted (", na_percent, "% NA)\n")
  }
} else {
  cat("FAILED - Test not completed\n")
}

# Test 8: Multiple files, both metadata
cat("Test 8 (Multiple files, both metadata): ")
if (exists("result_multiple_both") && "carat" %in% names(result_multiple_both)) {
  na_percent <- round(sum(is.na(result_multiple_both$carat))/nrow(result_multiple_both) * 100, 1)
  if (na_percent < 10) {
    cat("PASSED - Carat data preserved (", na_percent, "% NA)\n")
  } else {
    cat("FAILED - Carat data corrupted (", na_percent, "% NA)\n")
  }
} else {
  cat("FAILED - Test not completed\n")
}
```

## Package Status

### **Issues Resolved:**
1. **Metadata Parsing Conflicts**: Fixed the complex parsing logic that was corrupting CSV data
2. **Multiple Files Handling**: Resolved issues with `carat` variable becoming NA
3. **Cross-Platform Compatibility**: Package works on Windows, macOS, and Linux
4. **Relative Paths**: All hardcoded paths removed for worldwide distribution
5. **Data.table Bindings**: All global variable bindings properly declared